On the Fourier Analysis of Boolean Functions and the Linai-Mansour-Nisan Theorem

### 2020-11-09 CS590/project/paper.tex
Added FT notes to CS590

@@

Let us now consider the simplest finite abelian group, $\mathbb{Z}_n$ for some $n \in \mathbb{N}$. It follows from basic properties of cyclic groups that the only characters of $\mathbb{Z}_n$ are:
%
\begin{equation}
  \chi_j(x) = e^{ 2\pi i jx/n} \quad j \in [n], \; x \in \mathbb{Z}_n
\end{equation}
%
We can decompose any finite abelian group $G$ into a direct product of finite cyclic groups $G \cong \mathbb{Z}_{n_1} \times \mathbb{Z}_{n_2} \cdots \times\mathbb{Z}_{n_k}$. If we decompose $x = \sum_{i \in [k]} x_i$ for $x \in G$, then the characters of $G$ are defined as
%
\begin{equation}
  \chi_a(x) = \prod_{i \in [k]} e^{2 \pi i a_k x_k/ n_k}
\end{equation}
%
We see that every element $a \in G$ has an unique associated character $\chi_a$. Furthermore
%
\begin{gather} \label{chargroup}
  \chi_{a+b}(x)  = \chi_{a}(x)\chi_{b}(x) \\
  \chi_{a^{-1}}(x) = \frac{1}{\chi_a(x)}
\end{gather}
%
This shows that the set whose elements are the characters of $G$ endowed with pointwise multiplication is an abelian group. Denote this set as $\hat{G}$. From the map $a \mapsto \chi_a$, we have a group isomorphism between $G \cong \hat{G}$. The group $\hat{G}$ is known as the \emph{character group} of $G$. \newline

To define the Fourier transform, we must consider the Hilbert space $L_2(G)$ where $G$ endowed with the uniform probability measurel i.e the discrete measure mapping each subset $H \subseteq G$ to $\frac{|H|}{|G|}$. This yields the standard inner product for maps $f,g \in L_2(G)$:
%
\begin{equation}
  \langle f,g \rangle = \frac{1}{|G|}\sum_{x\in G} f(x)\overline{g(x)}= \mathbb{E}_x f(x)\overline{g(x)}
\end{equation}
Note that this induces the norm $||f||_2 = \sqrt{\langle f, f \rangle} = \mathbb{E}_x |f(x)|^2$ on $L_2(G)$. It is simple to check that the characters of $G$ indeed lie in $L_2(G)$. However, we can prove something stronger: the characters of $G$, $\chi_a$ form an \emph{orthonormal basis} of $L_2(G)$.
%

\begin{theorem}
  The characters $\chi_a \in \hat{G}$ form an orthonormal basis for $L_2(G)$
\end{theorem}

\begin{proof}
  To show this, we must employ a lemma:

\end{proof}

We restrict our attention to the finite abelian group $\mathbb{Z}_2^n$ as a natural group to define boolean functions $f: \{0,1\}^n \rightarrow \{0,1\}$. By the observations above, we conclude that our characters for $\mathbb{Z}_2^n$ are defined as:

\begin{equation}
  \chi_a(x) = (-1)^{\sum_{i \in [n]} x_ia_i} = (-1)^{\sum_{i \in [n], \; a_i = 1} x_i}
\end{equation}
where $x_i$ is the $i^{th}$ bit of $x \in \{0,1\}^n$.

>>>>

\theoremstyle{remark}
\newtheorem{notation}[theorem]{Notation}

@@

Consider the set whose elements are the characters of $G$ and whose binary operation is pointwise multiplication of complex-valued functions on $G$. This set is also an \emph{abelian group} by the following theorem:
%
\begin{theorem}
  If $G$ is a finite abelian group, then the characters of $G$ form an abelian group under pointwise multiplication.
\end{theorem}
%
\begin{proof}
  We first show that if $\chi, \phi$ are characters of $G$, then $\chi\phi$ is also a character of $G$. But this immediately follows from \ref{homo}:
  $$ \chi\phi(a+b) = \chi(a)\chi(b)\phi(a)\phi(b) = [\chi(a)\phi(a)][\chi(b)\phi(b)] = \chi\phi(a)\chi\phi(b)$$
  Hence, $\chi\phi$ is also a character of $G$. Furthermore, $\chi^{-1} = \frac{1}{\chi} = \overline{\chi}$ is also a character of $G$ since
  $$ \chi^{-1}(a+b) = \frac{1}{\chi(a+b)} = \frac{1}{\chi(a)\chi(b)} = \frac{1}{\chi(a)}\frac{1}{\chi(b)} = \chi^{-1}(a)\chi^{-1}(b)$$
  showing that the set has inverses.
\end{proof}
%
\begin{remark}
  We abuse notation a bit and just note that $\chi^{-1}$ is not the inverse of the $\chi$ as a \emph{map}. Rather, $\chi^{-1}$ will refer to the \emph{group inverse} of character $\chi$ from hereon.
\end{remark}

%
Denote this set as $\hat{G}$. The group $\hat{G}$ is known as the \emph{character group} of $G$. It also will turn out that there is a bijective map between $G$ and $\hat{G}$ such that the map is also a \emph{group isomorphism}. We will prove this later on. \newline

Let us now consider the simplest finite abelian group, $\mathbb{Z}_n = \mathbb{Z}/n\mathbb{Z}$ for some $n \in \mathbb{N}$. It follows from basic properties of cyclic groups that the only characters of $\mathbb{Z}_n$ are:
%
\begin{equation}
  \chi_j(x) = e^{ 2\pi i jx/n} \quad j \in [n], \; x \in \mathbb{Z}_n
\end{equation}
%
We can decompose any finite abelian group $G$ into a direct product of finite cyclic groups $G \cong \mathbb{Z}_{n_1} \times \mathbb{Z}_{n_2} \cdots \times\mathbb{Z}_{n_k}$. If we decompose $x = \sum_{i \in [k]} x_i$ for $x \in G$, then the characters of $G$ are defined as
%
\begin{equation}
  \chi_a(x) = \prod_{i \in [k]} e^{2 \pi i a_k x_k/ n_k}
\end{equation}
where $x_i$ refers to the $i^{th}$ index of $x$. \newline

We see that every element $a \in G$ has an unique associated character $\chi_a$. Furthermore
%
\begin{gather} \label{chargroup}
  \chi_{a+b}(x)  = \chi_{a}(x)\chi_{b}(x)
\end{gather} \newline

To define the Fourier transform, we must consider the Hilbert space $L_2(G)$ where $G$ endowed with the uniform probability measure i.e the discrete measure mapping each subset $H \subseteq G$ to $\frac{|H|}{|G|}$. This yields the standard inner product for maps $f,g \in L_2(G)$:
%
\begin{equation}
  \langle f,g \rangle = \frac{1}{|G|}\sum_{x\in G} f(x)\overline{g(x)}= \mathbb{E}_x f(x)\overline{g(x)}
\end{equation}
Note that this induces the norm $||f||_2^2 = \langle f, f \rangle = \mathbb{E}_x |f(x)|^2$ on $L_2(G)$. It is simple to check that the characters of $G$ indeed lie in $L_2(G)$. However, we can prove something stronger: the characters of $G$, $\chi_a$ form an \emph{orthonormal basis} of $L_2(G)$.
%
First, we prove a lemma:
\begin{lemma} \label{totalsum}
  Let $\chi$ be a non-trivial character of $G$. Then $\sum_{x \in G} \chi(x) = 0$
\end{lemma}
%
\begin{proof}
  Given any non-trivial character $\chi$, the following holds for all $y \in G$:
    %
    \[ \chi(y)\sum_{x \in G} \chi(x) = \sum_{x \in G} \chi(y+x) = \sum_{x \in G} \chi(x) \]
    %
    If $\sum_{x \in G} \chi(x) \neq 0$, then this would force $\chi(y) = 1$ for all $y \in G$. However, this would contradict the non-triviality of the $\chi$.
\end{proof}

\begin{theorem} \label{orthonormal}
  The characters $\chi \in \hat{G}$ form an orthonormal basis for $L_2(G)$
\end{theorem}
%
\begin{proof}
We first note that $\langle \chi, \chi \rangle = \mathbb{E}_x|\chi(x)|^2 = 1$ since the image of character $\chi$ lies on the complex unit circle by the discussions above. Now given two non-trivial characters $\chi, \phi$
$$ \langle \chi, \phi \rangle = \mathbb{E}_x \chi(x)\overline{\phi(x)} = \frac{1}{|G|} \sum_{x \in G} \chi\phi^{-1}(x)  = 0$$
The last equality follows from Lemma \ref{totalsum} since $\chi\phi^{-1}$ must also be a non-trivial character if $\chi, \phi$ are non-trivial characters. The case where one of the character is trivial also follows from the lemma.
Finally, observe that $L_2(G)$ is $|G|$-dimensional as a vector space by considering the dual of $G$. We know that there are at least $|G|$-characters, showing that the characters span $L_2(G)$. However, from the lemma, the characters must also be pairwise orthgonal to each other. This yields that the characters of $G$ form an \emph{orthonormal basis} of $L_2(G)$ and that there must be exactly $|G|$ characters in $\hat{G}$. Finally, by using \ref{chargroup}, it follows that the map $a \mapsto \chi_a$ must be a group isomorphism, so $G \cong \hat{G}$.
\end{proof}

To finish, we restrict our attention to the finite abelian group $\mathbb{Z}_2^n$ as a natural group to define boolean functions $f: \{0,1\}^n \rightarrow \{0,1\}$. By the observations above, we conclude that our characters for $\mathbb{Z}_2^n$ are defined as:
%
\begin{equation}
  \chi_a(x) = (-1)^{\sum_{i \in [n]} x_ia_i} = (-1)^{\sum_{i \in [n], \; a_i = 1} x_i}
\end{equation}
where $x_i$ is the $i^{th}$ bit of $x \in \{0,1\}^n$.
%
\begin{notation}
  Instead of considering $x \in \{0,1\}^n$ as a bit-string, we would like to use the following notation: Let $S \subseteq [n]$ be such that $S_x = \{i \mid x_i = 1\}$. As there is a bijective correspondence between all such subsets $S_x$ and $x \in \{0,1\}^n$, we will start to identify bit strings with their subset counterparts for notational convenience.
\end{notation}

\subsection{The Fourier Transform}

We have shown above that the members of the character group $\hat{G}$ form an orthonormal basis of $L_2(G)$. This bestows us the power of decomposing any $f \in L_2(G)$ into its orthogonal counterparts. The restriction to the case where $G = \mathbb{Z}_2^n$ is what motivates us as any $n$-ary boolean function $f: \{0,1\}^n \rightarrow \{0,1\}$ must lie in $L_2(\mathbb{Z}_2^n)$. As before, we begin by considering the general case for any finite abelian group $G$  \newline

Let $\hat{f}: \hat{G} \rightarrow \mathbb{C}$ be the complex-valued function defined as:
%
\begin{equation}
  \hat{f}(\chi_a) =  \langle f, \chi_a \rangle, \; a \in G
\end{equation}
%
We can iterate this for all characters to get a direct sum decomposition for any $f \in L_2(G)$
%
\begin{equation}
  f = \sum_{a \in G} \hat{f}(\chi_a)\chi_a
\end{equation}

It can be show that the map $f \mapsto \hat{f}$ is linear by the left-linearity of the inner product. This linear map is the \emph{Fourier Transform on $G$}.
%
\begin{notation}
  We will sometimes denote $\hat{f}(a) = \hat{f}(\chi_a)$.
\end{notation}

A simple application of the Fourier Transform yields an useful identity which will be instrumental in our formulation of the Fourier tail of a boolean function.

\begin{theorem}{(Parseval's Identity)}
  Let $f \in L_2(G)$. Then
  $$ ||f||_2^2 = \sum_{a \in G} |\hat{f}(a)|^2$$
\end{theorem}
%
\begin{proof}
  $$ ||f||^2_2 = \langle f, f \rangle = \langle \sum_{a \in G} \hat{f}(a) \chi_a,  \sum_{b \in G} \hat{f}(b) \chi_b \rangle = \sum_{a,b \in G} \hat{f}(a)\overline{\hat{f}(b)}\langle \chi_a, \chi_b \rangle = \sum_{a \in G} |\hat{f}(a)|^2 $$
  The last equality is just the orthonormality of the characters (\ref{orthonormal}).
\end{proof}

%by the remark in the previous section, we will identify characters of %$\mathbb{Z}_2^n$ as $\chi_A$ for $A \subseteq [n]$.
